A Plasma Torus Around a Young Low-Mass Star

### 2025-03-19 paper/ms.tex
update

The physical conditions inside the circumstellar plasma clumps, in
particular the hydrogen number density, plasma temperature, ionization
fraction, and magnetic field strength can all be estimated using the
available data.  We caution that the following estimates are coarse:
detailed considerations of radiative transfer and detailed balance are
a worthwhile topic for future work, but are beyond the scope of this
article.

Circumstellar H$\alpha$ emission might be sourced either from
scattering of stellar H$\alpha$ photons, or from recombination.  We
neglect scattering because we see the amplitude of the circumstellar
emission to vary by a factor of $\approx$5, which is not mirrored in
the chromospheric line core.  The volume emissivity under case B
recombination (in which Lyman-series photons are trapped by high
optical depth, so that recombinations cascade through H$\alpha$ and
other Balmer lines) can be written as
\begin{equation}
  j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha},
\end{equation}
where $n_{\rm e}$ and $n_{\rm p}$ are the electron and proton densities,
and $\alpha^{\rm eff}_{\rm H\alpha}(T)$ is the effective recombination
coefficient, i.e.\ the number of recombinations per unit volume, per
unit time.
For hydrogen with temperatures between 1,000 and 10,000\,K with
densities $n_{\rm H}$$\sim$10$^{10}$\,cm$^{-3}$,
$\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the order of $10^{-12}$ to
$10^{-13}$\,cm$^3$\,s$^{-1}$ \cite{Hummer1987}.
Neglecting the effects of atoms other than hydrogen, we can assume
an ionization fraction $x$, such that $n_{\rm e} = n_{\rm p} = x
n_{\rm H}$, for $n_{\rm H}$ the hydrogen number density.  Let $L_{\rm
H\alpha} = j_{\rm H\alpha} V$, for $V$ the volume of the emitting
hydrogen.
The luminosity of circumstellar hydrogen emission, $L_{\rm H\alpha}$,
is an observable: our SED fitting routine yields
$L_\star$$\approx$0.012\,$L_\odot$, which gives
$L_{\rm H\alpha},\star$$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$, and
the luminosity of the clump $L_{\rm H\alpha}$ is of order one tenth
that of the star.
Assuming an emitting sphere of radius $r$, we can write
\begin{equation}
  n_H = \frac{1}{x}
  \left( 
    \frac{L_{\rm H\alpha}}{ \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha} V }
  \right)^{1/2}.
\end{equation}
Taking nominal parameters of $r = 0.1 R_\star$, $x=1/2$, and
$\alpha^{\rm eff}_{\rm H\alpha}$=$10^{-13}$\,cm$^3$\,s$^{-1}$,
we find $n_{\rm H}\approx 5\times10^{11}$\,cm$^{-3}$.
Assuming a uniform density clump, this yields $M_{\rm gas}\approx
8\times10^{16}$\,g.

Finally, a constraint on the magnetic field strength at the site of the clump
follows from the requirement that the magnetic pressure exceed the
thermal pressure, $B^2 / 8\pi > n_{\rm H} k T$.
Although we do not know the plasma temperature,
if it were significantly outside the range of
1,000 and 10,000\,K,
we would either full ionize the hydrogen, or not ionized enough of it.
We can therefore estimate
\begin{equation}
  B \gtrsim 3\,{\rm G}
  \left(
  \frac{n_{\rm H}}{5\times10^{11}\,{\rm cm}^{-3}}
  \frac{T}{5000\,{\rm K}}
  \right).
\end{equation}
Given that the average surface magnetic field strengths of low-mass
stars have been measured to span hundreds to thousands of Gauss
\cite{Kochukhov2021}, this bound is easily met.

@@

Broadly, our calculations suggest that if dust is responsible for the
broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.  Given
$M_{\rm gas}$$\approx$8$\times$10$^{16}$\,g from
Section~\ref{subsec:gas}, this range spans mass ratio ranges $M_{\rm
gas}/M_{\rm dust}$ of 100 to 1.  More careful measurements of this
ratio---in particular enabled by inferring the dust mass through high
precision infrared spectrophotometry---would provide a pathway for
distinguishing between environments that more closely resemble the
interstellar medium versus protoplanetary disks.

@@

\item[Acknowledgments]
  % convos w/
  % M.~Jardine. (line core sum, explaining flux tubes)
  % A Weinberger (dissolving rocks)
  % L Hillenbrand (general?)
  % line profile analysis by B.~Tofflemire.
  % CPS group for reduction support.
  The author thanks X, Y, Z.
  L.G.B. was suported by...
	Acknowledge TESS...
  Acknowledge use of IRSA.

%TC:ignore
%% Author Contribution
\item[Author Contributions] ...
%TC:endignore

\item[Data Availability] ...

\item[Competing Interests] The authors declare that they have no competing
financial interests.

\item[Correspondence] Correspondence and requests for materials should be
addressed to ...

>>>>

The physical conditions inside the plasma clumps, in particular the
hydrogen number density, plasma temperature, ionization fraction, and
magnetic field strength can all be estimated using the available data.
We caution however that the following estimates are coarse: detailed
considerations of radiative transfer and detailed balance are a
worthwhile topic for future work, but are beyond the scope of this
article.

Circumstellar H$\alpha$ emission might be sourced either from
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figures~\ref{fig:spec}
and~\ref{fig:hirescuts} show the amplitude of the circumstellar
H$\alpha$ emission varying by a factor of $\approx$5, in a manner
uncorrelated with any variability in the chromospheric line core.  The
volume emissivity under case B recombination (in which Lyman-series
photons are trapped by high optical depth, so that recombinations
cascade through H$\alpha$ and other Balmer lines) can be written as
\begin{equation}
  j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha},
\end{equation}
where $n_{\rm e}$ and $n_{\rm p}$ are the electron and proton densities,
and $\alpha^{\rm eff}_{\rm H\alpha}(T)$ is the effective recombination
coefficient, i.e.\ the number of recombinations per unit volume, per
unit time.
For hydrogen with temperatures between 1,000 and 10,000\,K with
densities $n_{\rm H}$$\sim$10$^{10}$\,cm$^{-3}$,
$\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the order of $10^{-12}$ to
$10^{-13}$\,cm$^3$\,s$^{-1}$ \cite{Hummer1987}.
Neglecting the effects of atoms other than hydrogen, we can assume
an ionization fraction $x$, such that $n_{\rm e} = n_{\rm p} = x
n_{\rm H}$, for $n_{\rm H}$ the hydrogen number density.  Let $L_{\rm
H\alpha} = j_{\rm H\alpha} V$, for $V$ the volume of the emitting
hydrogen.
The luminosity of circumstellar hydrogen emission, $L_{\rm H\alpha}$,
is an observable: our SED fitting routine yields
$L_\star$$\approx$0.012\,$L_\odot$, which gives
$L_{\rm H\alpha},\star$$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$, and
the luminosity of the clump $L_{\rm H\alpha}$ is of order one tenth
that of the star.
Assuming an emitting sphere of radius $r$, we can write
\begin{equation}
  n_H = \frac{1}{x}
  \left( 
    \frac{L_{\rm H\alpha}}{ \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha} V }
  \right)^{1/2}.
\end{equation}
Taking nominal parameters of $r = 0.1 R_\star$, $x=1/2$, and
$\alpha^{\rm eff}_{\rm H\alpha}$=$10^{-13}$\,cm$^3$\,s$^{-1}$,
we find $n_{\rm H}\approx 5\times10^{11}$\,cm$^{-3}$.
Assuming a uniform density clump, this yields a total gas mass $M_{\rm gas}\approx
8\times10^{16}$\,g.

Finally, a constraint on the magnetic field strength at the site of the clump
follows from the requirement that the magnetic pressure exceed the
thermal pressure, $B^2 / 8\pi > n_{\rm H} k T$.
Although we do not know the plasma temperature,
if it were significantly beyond 1,000-10,000\,K,
we would either full ionize the hydrogen, or not ionized enough of it.
We can therefore estimate
\begin{equation}
  B \gtrsim 3\,{\rm G}
  \left(
  \frac{n_{\rm H}}{5\times10^{11}\,{\rm cm}^{-3}}
  \frac{T}{5000\,{\rm K}}
  \right).
\end{equation}
Given that the average surface magnetic field strengths of low-mass
stars have been measured to span hundreds to thousands of Gauss
\cite{Kochukhov2021}, this bound is easily met.

@@

Broadly, our calculations suggest that if dust is responsible for the
broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.  Given
$M_{\rm gas}$$\approx$8$\times$10$^{16}$\,g from
Section~\ref{subsec:gas}, this range spans mass ratio ranges $M_{\rm
gas}/M_{\rm dust}$ of 100 to 1.  More careful measurements of this
ratio---in particular enabled by inferring the dust mass through high
precision infrared spectrophotometry---could provide a pathway for
distinguishing the scenario of a stellar outflow from an 
accumulation of extrinsically-sourced material.

@@

\item[Acknowledgments]
  LGB is grateful to M.~Jardine for insightful conversations,
  A.~Weinberger for clarification of infrared dust spectroscopy,
  B.~Tofflemire for a clarifying line profile analysis, and to both
  A.~Howard and H.~Isaacson for their assistance reducing the HIRES
  spectra.
  LGB was supported by the Carnegie Institution for Science's Carnegie
  Fellowship and the Heising-Simons 51~Pegasi~b Fellowship.
  Funding for the TESS mission is provided by NASA’s
  Science Mission directorate.
  The TESS mission is a product of millions of hours of work by
  thousands of people, and we thank the many TESS team members for
  their efforts to make the mission a continued success.
  The HIRES data were obtained at the Keck Observatory, which exists
  through a similar scale of community effort.
  We recognize the importance that the summit of Maunakea has always had
  within the indigenous Hawaiian community, and we are deeply grateful 
  for the opportunity to conduct observations from this mountain.

\item[Author Contributions] LGB is responsible for all aspects of this
  work, including conceptualization, data analysis, interpretation,
  manuscript preparation, and revision.

\item[Data Availability] The TESS two-minute cadence data used in this
  work are publicly available from the Mikulski Archive for Space
  Telescopes (MAST; \url{https://mast.stsci.edu}). The Keck/HIRES
  spectra analyzed here are accessible via the Keck Observatory
  Archive (KOA; \url{https://koa.ipac.caltech.edu}).

\item[Competing Interests] The author declares no competing financial
  interests.

\item[Correspondence] Correspondence and requests for materials should
  be addressed to Luke Bouma, at either lbouma@carnegiescience.edu or
  bouma.luke@gmail.com.
